---
output:
  md_document:
    variant: gfm
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# abmidata

> Public data from the Alberta Biodiversity Monitoring Institute

[![check](https://github.com/ABbiodiversity/abmidata/workflows/check/badge.svg)](analythium/covid-19)

## Setup

Install the package:

```{r eval=FALSE}
remotes::install_github("ABbiodiversity/abmidata")
```

## Usage

Load the package

```{r}
library(abmidata)
```

List table names

```{r}
n <- ad_get_table_names()
data.frame(Description = head(n))
```

Get a table

```{r}
x <- ad_get_table("T01A", year=2010)
str(x)
```

Handling missing value indicators

```{r}
y <- x[["Aspect (degrees)"]][1:100]
y
ad_convert_na(y)
summary(ad_process_na(y))
```

## Documentation

See the [package website](https://abbiodiversity.github.io/abmidata) for documentation.
A list of tables and associated metadata can be found [here](https://abbiodiversity.github.io/abmidata/articles/abmidata02-metadata.html).
Field protocols and the table metadata can be found
in [this zip file](https://github.com/ABbiodiversity/abmidata/raw/master/metadata/DESCRIPTIONDATA.zip) as well.

## Issues

[https://github.com/ABbiodiversity/abmidata/issues](https://github.com/ABbiodiversity/abmidata/issues)

## Citation

```{r results='asis',echo=FALSE}
print(citation("abmidata"), style = "text")
```

## License

[MIT](LICENSE) © 2020 Peter Solymos, Joan Fang, ABMI
