---
output:
  md_document:
    variant: gfm
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# abmidata

> Public data from the Alberta Biodiversity Monitoring Institute

[![check](https://github.com/ABbiodiversity/abmidata/workflows/check/badge.svg)](https://github.com/ABbiodiversity/abmidata/actions)

## Setup

Install the package:

```{r eval=FALSE}
remotes::install_github("ABbiodiversity/abmidata")
```

## Usage

### Pulling monitoring data

The ABMI public data API pulls information from one of ABMI's main databases.

Load the R package:

```{r}
library(abmidata)
```

List names of tables available via the API:

```{r}
n <- ad_get_table_names()
data.frame(Description = head(n))
```

Get a table by its ID, possibly filtered by site ID or by year:

```{r}
x <- ad_get_table("T01A", year=2010)
str(x)
```

ABMI data has specific placeholders for missing data indicating
the [reason for the missingness](https://abbiodiversity.github.io/abmidata/reference/helpers.html). This poses challenges in R when
the information is numeric. Here is how you can handle
these missing value indicators:

```{r}
y <- x[["Aspect (degrees)"]][1:100]
y
ad_convert_na(y)
summary(ad_process_na(y))
```

### Processing data

Once tables are loaded using the ABMI public data API, the data
usually needs to be processed further.
The tables are in 'long format' which means that e.g. sites and species
are all in in their own columns:

| Site   | Species | Count |
|--------|---------|-------|
| Site 1 | A       | 1     |
| Site 1 | B       | 2     |
| Site 1 | C       | 3     |
| Site 2 | B       | 1     |
| Site 2 | D       | 2     |
| Site 3 | E       | 1     |

There are other R packages that can be used to turn the 'long format'
data into a 'wide format' where e.g. sites represent rows and species
represent columns:

| Site | A | B | C | D | E |
|------|---|---|---|---|---|
| 1    | 1 | 2 | 3 | 0 | 0 |
| 2    | 0 | 1 | 0 | 2 | 0 |
| 3    | 0 | 0 | 0 | 0 | 1 |

Check out examples of how to do this
for vascular plants, mites, mosses, lichens, and birds
in [this article](https://abbiodiversity.github.io/abmidata/articles/abmidata03-commontasks.html) about common data processing tasks.

### Matching monitoring data with predictors

Once the monitoring data is in the desired format, one can match the site locations
(and possibly years) with location/year specific predictor variables.
One can use the **field data** about site condition to explain the distribution,
abundance, or other measures of species and habitat elements at the ABMI sites.

One can also use **geospatial information** extracted at the site coordinates.
This, however, poses some challenges because exact site coordinates are
not publicly available. However, restrictions around site confidentiality
have been substantially relaxed and coordinates for most sites will soon be 
available for research purposes upon request. We will provide relevant information
here as soon as it is available. In the meantime, please
reach out to [ABMI](https://www.abmi.ca/home/contact-us.html) for more information.

## Documentation

See the [package website](https://abbiodiversity.github.io/abmidata) for documentation.
A list of tables and associated metadata can be found [here](https://abbiodiversity.github.io/abmidata/articles/abmidata02-metadata.html).

See also:

- [Terrestrial Field Data Collection Protocols](https://abmi.ca/home/publications/501-550/549)
- [Wetland Field Data Collection Protocols](https://abmi.ca/home/publications/501-550/548)
- [Terrestrial ABMI Autonomous Recording Unit (ARU) and Remote Camera Trap Protocols](https://abmi.ca/home/publications/551-600/565)

## Issues

[https://github.com/ABbiodiversity/abmidata/issues](https://github.com/ABbiodiversity/abmidata/issues)

## Citation

```{r results='asis',echo=FALSE}
print(citation("abmidata"), style = "text")
```

## License

[MIT](LICENSE) © 2020 Peter Solymos, Joan Fang, ABMI
